Canopy leaf area of an evergreen Eucalyptus woodland does not respond to elevated

### 2015-05-08 manuscript.Rmd
Manuscript updates; Bug fix (Hmisc needed after all); airt includes more met data.

Leaf area index (*L*) is an important ecosystem attribute underlying productivity of vegetation [@Reich2012], and a key variable in land surface models [@Kala2013; @Walker2014]. *L* is a dynamic canopy property, as it is generally responsive to variation in water and nutrient availability, as demonstrated in experiments [@Herbert1995; @Hubbard2004; @White2010] and observational studies across rainfall gradients [@Ellis2008; @Iio2013; @Jin2013].  Ecosystem models generally predict that *L* increases in response to  due to increased net carbon uptake [@DeKauwe2014; @Hatton1992; @Kergoat2002; @Zhu2012]. Other mechanisms are likely involved as well, including increased quantum yield of photosynthesis in eC~a~ allowing leaf retention into deeper shade [@Norby1999], and interactions with improved water relations under eC~a~  [@Wullschleger2002]. Quantifying the magnitude and direction of the effect of elevated atmospheric [CO~2~] (eC~a~) on *L* is crucial, as changes in *L* associated with climate change drivers such as increases in C~a~ may provide an important feedback to climate change, and modify direct physiological responses to C~a~ [@Tor-ngern2015; @Murray2012]. To date, little empirical evidence is available on the response of *L* to eC~a~ in mature forests [@Norby2011].

\
Direct effects of eC~a~ on vegetation processes include an enhancement of photosynthesis and productivity, a decrease in stomatal conductance and leaf-level water use and an increase in the water-use efficiency [@Ainsworth2005; @Ellsworth2012]. Although numerous studies have been conducted on the response of physiology and plant growth to eC~a~, very few of these inform the response of steady-state *L* to eC~a~, because it requires a sufficiently mature forest to be exposed to free air CO~2~ enrichment (FACE) [@Norby2011]. Increased carbon availability through enhanced photosynthesis under eC~a~ does not necessarily lead to increases in *L*. Out of the six forest FACE studies that have been carried out worldwide, only broad-leaved deciduous aspen stands (in monoculture or mixed with birch) and needle-leaved evergreen pine stands showed increased *L* in response to eC~a~ *L* [@Norby2011], although positive responses to eC~a~ in terms of productivity were observed at all six sites. Interestingly, Norby and Zak [-@Norby2011] found a negative correlation between pre-treatment *L* and the responsiveness to eC~a~, leading to the prediction that stands with low *L* will show a larger response to eC~a~.

\
In ecosystems where water availability limits productivity, we could expect eC~a~ to further affect *L* [@Donohue2009], since eC~a~ reduces stomatal conductance, and hence water use per unit of leaf area [@Warren2010] leading to increased soil water availability. These ‘water savings’ can be expected to lead to an increase in *L* in water-limited ecosystems, since we observe a tight coupling of *L* with precipitation across gradients [@Ma2013; @Jin2013; @Hatton2008]. In support of this prediction, Donohue et al. [-@Donohue2013] attributed the observed global increase in *L* in arid and semi-arid areas over the last 30 years to rising C~a~. Based on the same water-savings principle, we predict that *L* should increase under eC~a~ in water-limited forests.

\
Most previous studies on *L* responses to environmental drivers have focused on the seasonal, mostly temperature and light-driven, timing of *L* dynamics observed in N-hemisphere forests [e.g. @Richardson2012; @Savoy2015] whereas *L* responses to fluctuations in water availability are qualitatively well-known [@Pook1985; @Specht1989; @Whitehead2004; @Pook1997; @Limousin2009], but poorly quantified. The strong relationship between *L* and rainfall observed across Australia [@Ellis2008; @Specht1989; @Broich2014], and other continents [@Donohue2013; @Grier1977; @Jin2013] represents the adaptation of vegetation structure and density to long-term climate conditions. However, local *L* is also highly responsive to water availability at shorter time scales [@Pook1985; @Pook1997], driven by leaf turnover [@Misson2011] and leaf flushing and expansion [@Muller2011], which are both highly responsive to soil water availability. Nevertheless, few datasets are currently available at sufficiently high temporal resolution to allow modelling of water-driven changes in phenology [@Misson2011].

\
We estimated *L* from near-continuous measurements of canopy transmittance in a *Eucalyptus*-dominated woodland exposed to eC~a~, the EucFACE experiment. The woodland at the EucFACE is frequently water-limited, allowing us to quantify the dynamics of *L*, and its components (litter fall and production) in response to water availability. We hypothesized that eC~a~ would result in an increase in *L*, particularly during periods of water limitation, as water savings were expected to lead to longer retention of foliage under elevated CO~2~. In addition, we expected that the low *L* that is typical of native *Eucalyptus*-dominated woodlands [@Whitehead2004] allowed a larger response to eC~a~ [@Norby2011]. Our second objective was to quantify the temporal dynamics of *L*, in particular in relation to fluctuating water availability.

@@

## Site description and experimental design
The EucFACE experiment is located on an ancient alluvial floodplain about 5 km from the Hawkesbury River in western Sydney (Australia) at 30 m a.s.l. elevation (33°37'S, 150°44'E). It is located within a remnant patch of native Cumberland Plain woodland (Tozer, 2003). The vegetation, maintained in this location since the 1880’s, retains some old-growth trees and some re-growth vegetation though it has been undisturbed for ca. 70 years. The site is characterized by a humid temperate-subtropical transitional climate with a mean annual temperature of 17 °C and mean annual precipitation of 800 mm (Bureau of Meteorology, station 067105 in Richmond, NSW Australia; http://www.bom.gov.au).

\
The soil at the site is a loamy sand in the Clarendon Formation (Barton et al., 2010; Isbell, 2002), a slightly-acid, low-organic C, well-drained soil. The soil surface is a loamy sand with > 75% sand content, and from ca. 50 to 300 cm depth, soils are a sandy clay loam with > 30% silt and clay. The soil is nutrient poor, and a strong limitation of P on tree growth at the site has recently been demonstrated [@Crous2015]. Groundwater depth is 12.6 – 12.9 m below the soil surface. The woodland ranges in density from 600-1000 trees ha^-1^. The vegetation consists of an open woodland (basal area before treatment, BA = `r mean(ba$BA)`, SE = `r sd(ba$BA)/sqrt(6)` m^2^ha^-1^, n = 6 plots) with a single major canopy-forming tree (*Eucalyptus tereticornis* Sm.) with a minor component of *E. amplifolia* Naudin. The understory stratum is composed mainly of grass species dominated by *Microlaena stipoides* (Labill.) R.Br., with lower densities of forbs of *Commelina cyanea* R.Br., *Lobelia purpurascens* R.Br., *Dichondra repens* J.R.Forst. & G.Forst and occasional shrubs of *Breynia oblongifolia* (Mull.Arg.) Mull.Arg. and *Bursaria spinosa* Cav.

@@

where $\tau_p$ canopy transmission estimated from canopy photos. We assumed that the extinction coefficient *k* = 0.5, following Macfarlane et al. [-@Macfarlane2007] recommendation for eucalypt forests. Without further calibration, the two estimates were reasonably well correlated, though on average $L_\tau$ was `r 100*(with(flatcan_byring, mean(LAI.PAR.mean / LAI))-1)` higher than $L_p$ (Fig. SI2, $R^2$=`r glance(lmphoto)$r.squared`, RMSE=`r glance(lmphoto)$sigma`).

, and compared this assumption with an optimized extinction that minimized the difference between the two estimates of *L*.

\
We compared $L_\tau$ (after calibration to litter production) to the independent estimate of *L* from canopy photos ($L_p$).  This is encouraging because the photos measure *L* near the zenith (+/- 15°), while $L_\tau$ is integrated over the sky hemisphere (Eq. 3), weighted towards the zenith.

@@

## Response of *L* to elevated [CO~2~]

** tie to objectives at end of intro! **

We report on `r nummon` months of continuous leaf area index (*L*) measurements in a native evergreen *Eucalyptus* woodland exposed to +150ppm elevated CO~2~ (eC~a~) and 3 months of ramp-up in C~a~. We had predicted that *L* would increase in response to eC~a~ as a result of increased quantum yield for photosynthesis, enhanced CO~2~ assimilation for the entire canopy, and increased soil water availability. We also expected that this eucalypt woodland, with its inherently low *L* [@Whitehead2004], would be even more responsive to eC~a~ than higher-*L* and more closed-canopy stands [@Norby2011]. Counter to our expectations but similar to Norby et al. [-@Norby2003], we found no significant effect of elevated C~a~ on *L*. It could be argued that the change in *L* in response to eC~a~ was too small to be detected. Natural variation in *L* among plots in a native woodland is indeed a challenge to manipulative studies with low replication, such as FACE experiments. To increase sensitivity and statistical robustness, we used a novel method to account for pre-treatment differences by using stand basal area as a covariate. Using this method, we estimated that we can detect a difference of only a `r format(Ldetect,digits=1)`% in *L*. Based on this small detectable difference, we conclude that the lack of a significant C~a~ effect is not due to a lack of power, but a true lack of response to elevated C~a~.

\
This lack of a response to elevated C~a~ counters our expectations and those in previous publications [@Norby2003; @McCarthy2007], for a number of reasons. First, in a review of forest FACE sites, Norby and Zak [-@Norby2011] showed that low *L* stands were more responsive to eC~a~. They argued that high *L* canopies would not benefit from further increases in *L*, because it would lead to only very small increases in total absorbed radiation. The mean *L* at the EucFACE was ca. `r mean(ba$LAI.mean)` m^2^m^-2^, which is low compared to a global compendium of forest *L* [@Iio2013], and the lowest *L* of any forest FACE experiment [@Norby2011], but in line with a range of values for native *Eucalyptus* forests and woodlands reported by Whitehead and Beadle [-@Whitehead2004] and Iio et al. [-@Iio2013]. Based on this *L* estimate and the relationship across FACE sites reported by Norby and Zak [-@Norby2011], we could expect about a 30% increase in *L* under eC~a~. It is clearly too early to draw general conclusions across FACE sites in terms of *L* response, moreover because water and nutrient likely play a central role in productivity and *L* dynamics at the EucFACE. Indeed, experimental evidence clearly demonstrates strong limitations of *L* by nutrients and water in Eucalyptus stands [@White2010].

\
A possible explanation for the lack of a *L* response to eC~a~ is that leaf growth may not be directly related to increased carbon availability as a result of eC~a~. In other words, stimulation of net leaf photosynthesis by eC~a~ (a common response, also observed at the EucFACE, Gimeno et al. in review) may not lead to increased net leaf growth, due to other constraints such as water or nutrient limitations [@Crous2015]. Net leaf growth is the net sum of allocation of photosynthates to leaf growth, and turnover of the leaf pool. Either process may itself be affected by eC~a~ [@DeKauwe2014]. We found a significant effect of eC~a~ on leaf lifespan, with leaves living approximately `r round(abs(LLtest$estimate*12),1)` months longer in elevated C~a~ (Fig. 6). Because we found no difference in *L* with C~a~ treatment, this suggests a compensating difference in the fractional allocation of net primary production to leaf growth. To further understand the response of biomass allocation to eC~a~ at the EucFACE, all components of the carbon balance must be carefully quantified, a task underway but not yet complete. For example, Drake et al. (in review) reported increased soil respiration under eC~a~, suggesting that at least some of the stimulation of photosynthesis was allocated to either root growth, respiration, or root exudation.

\
Other studies in FACE are consistent with longer leaf-lifespan in elevated Ca shown here.
[@Taylor2008] reported delayed autumn senescence in *Populus* in two FACE experiments (PopFACE and AspenFACE). Although this indicates the canopy stayed greener longer in elevated C~a~, [@Gielen2001] reported faster turnover rate of leaves in elevated C~a~ due to overall higher productivity, leading to similar *L* in ambient C~a~ despite higher productivity.
But not with [@Warren2011]: increased senescence in eC~a~ in response to a combined drought and heat wave.

##Continuous measurement of *L*
Studying canopy leaf area index in an evergreen natural woodland poses unique challenges. Unlike in deciduous canopies, litter production cannot be used to estimate *L*, because leaf lifespan is unknown. We can also not rely on estimated leaf area based on allometric relationships with stem diameter or height, since these equations are not available for mature *E. tereticornis*, and besides these would be useless since *L* is highly variable temporally (Figs. 1, 4). We found that indirect estimates of *L* based on canopy transmittance were very reliable.  Estimated *L* compared well to that from flat canopy photos (cf. [@Macfarlane2007] (Fig. SI2), and after calibration, to leaf litter production during a dry period (Fig. 4). Unlike canopy photos, PAR sensors do not need thresholding or time-consuming post-processing. We do argue that the raw data must be filtered by diffuse conditions, as the spatial and within-day variance are negligible compared to periods of direct sunshine [@Chason1991; @Dufrene1995; @Nilson1999]. As expected, when we used unfiltered canopy transmittance data, the seasonal patterns in canopy transmittance were much more distinct (Fig. S3), because longer path lengths in the winter result in lower overall transmittance. Using diffuse PAR only avoids having to correct for these path length effects.

## Litter production and dynamics of *L*
We found a correlation between *L* dynamics and litter production, both during periods of decreasing and increasing *L*. During a four-month period in 2013 without significant rainfall, we used the decrease in *L* as an opportunity to calibrate against litter production. During periods of increasing *L*, litter fall increased sharply, indicating replacement of old leaves by young leaves. Indeed we observed in the field that after the significant flushing event of Dec. 2013, very few old leaves were present in the canopy. This is in line with [@Pook1984] in *Corymbia maculata* (a genus closely related to *Eucalyptus*), who showed accelerated litter production during periods of net *L* increase, but with a lag between litter fall and production of about 2 months. Similarly, [@Stone1995] showed that drought-stressed trees had higher litter production at the branch scale, but irrigated trees had larger litter production at the canopy scale. Since leaf flushing has been shown to be closely related to soil water availability [Fig. 7; @Pook1984a], net leaf turnover is driven by water availability both through increased litter production during extended droughts, and complete canopy replacement following leaf flushing (Fig. 5). The consequence is that leaf turnover is highly variable in time.

## Consequences for phenology models
Current-generation phenology models are heavily biased towards temperate conditions [@Primack2011], where leaf flushing occurs in spring in response to increase in temperature [@Richardson2012; @Savoy2015]. In contrast, phenology of evergreen *Eucalyptus* is more opportunistic, and shows rapid response to water availability (Fig. 7). Specht and Brouwer [-@Specht1975] demonstrated for four *Eucalyptus* species that leaf growth  peaked both in spring and autumn, with low winter temperatures and high summer temperatures (and low water availability) suppressing leaf growth. Similarly, Pook  [-@Pook1984a] demonstrated leaf production during most months in *Corymbia maculata*, except during winter and suppression by dry periods. Thus a minimum temperature is also required for leaf growth to occur. We also have not observed leaf flushing in winter time, but at our site this was confounded with water availability as both winters during the study period were relatively dry (Fig. 7).

\
Although the response of *L* to water availability is qualitatively well-known, current-generation phenology models fall short in incorporating this response. In a comparison of six leading ecosystem and global vegetation models,  Powell et al. [-@Powell2013] showed that the steady decrease in *L* in an Amazon drought experiment could be matched by only one model. Simulations from De Kauwe et al. [-@DeKauwe2014] at ambient and elevated C~a~ also showed little response to inter-annual variation in rainfall (pers. comm. Martin De Kauwe). In most models  that provide prognostic estimates of *L* [i.e., excluding models that take *L* as an input parameter; e.g. @Kala2013), temporal changes in *L* are modelled by first calculating total net photosynthetic productivity (carbon availability, gross photosynthesis minus respiration), and allocating some fraction to the leaf pool. Combined with - usually constant - leaf turnover and specific leaf area (ratio of leaf area to mass), this yields predictions of *L* in response to environmental drivers such as eC~a~ (Fig. 2). We hypothesize that this algorithm is not well suited to explain the dynamic nature of *L* in response to water availability observed in evergreen broad-leaved forests, because not all important mechanisms are incorporated in these models. In order to construct a phenology model where phenology is responsive to water availability, it should include responses of both leaf turnover and leaf growth to fluctuating water availability. Water limitation is known to directly influence leaf turnover [@Pook1997], possibly to maintain sufficient rates of water availability to the remaining leaves [@Vilagrosa2003; @Misson2011], and rates of leaf expansion [@Muller2011]. Some existing ecosystem models allow for increased leaf turnover during periods of low water availability (e.g. G’DAY, see Corbeels et al. 2005), but this process has been poorly quantified.

# Conclusions
Leaf area index (*L*) of a natural *Eucalyptus* woodland was unresponsive to elevated [CO~2~] after 27 months of treatment and 24 months at +150ppm CO~2~ elevation. Litter production was closely related to dynamics of *L*, both during dry periods (leading to accelerated litter production), and periods of net leaf growth (when old foliage was replaced by new foliage). Leaf lifespan, estimated from mean *L* and annual litter production, was `r round(abs(LLtest$estimate*12),1)` months longer in the elevated [CO~2~] treatment.

We found no clear annual pattern in dynamics of *L*, instead, a net increase in *L*, indicating leaf growth, consistently commenced after heavy rainfall following a dry period. The complex dynamics of litter production and the rapid response of litter production and leaf growth to soil water availability challenge current-generation prognostic *L* and phenology models. High resolution time series such as the one we present in this study can be used to develop new models of phenology for evergreen woodlands.

>>>>

Leaf area index (*L*) is an important ecosystem attribute underlying productivity of vegetation [@Reich2012], and a key variable in land surface models [@Kala2013; @Walker2014]. *L* is a dynamic canopy property, as it is generally responsive to variation in water and nutrient availability, as demonstrated in experiments [@Herbert1995; @Hubbard2004; @White2010] and observational studies across rainfall gradients [@Ellis2008; @Iio2013; @Jin2013].  Ecosystem models generally predict that *L* increases with eC~a~, largely in response to  due to increased net carbon uptake [@DeKauwe2014; @Hatton1992; @Kergoat2002; @Zhu2012]. This potential change in *L* is of central importance to net effect of eC~a~ on water, CO~2~ and energy fluxes as it may modify or even reverse the direct physiological responses to C~a~ [@Tor-ngern2015; @Murray2012]. Despite its importance, little empirical evidence is available on the response of *L* to eC~a~ in mature forests [@Norby2011].

\
Direct effects of eC~a~ on vegetation processes include an enhancement of photosynthesis and productivity, a decrease in stomatal conductance and leaf-level water use and an increase in the water-use efficiency [@Ainsworth2005; @Ellsworth2012]. Although numerous studies have been conducted on the response of physiology and plant growth to eC~a~, very few of these inform the response of steady-state *L* to eC~a~, because it requires a sufficiently mature forest to be exposed to free air CO~2~ enrichment (FACE) [@Norby2011]. Increased carbon availability through enhanced photosynthesis under eC~a~ does not necessarily lead to increases in *L*. Out of the six forest FACE studies that have been carried out worldwide, only broad-leaved deciduous aspen stands (in monoculture or mixed with birch) and needle-leaved evergreen pine stands showed increased *L* in response to eC~a~ *L* [@Norby2011], although positive responses to eC~a~ in terms of productivity were observed at all six sites [@Norby2005]. Interestingly, Norby and Zak [-@Norby2011] found a negative correlation between pre-treatment *L* and the responsiveness to eC~a~, leading to the prediction that stands with low *L* will show a larger response to eC~a~.

\
In ecosystems where water availability limits productivity, we could expect eC~a~ to further affect *L* [@Donohue2009], since eC~a~ reduces stomatal conductance, and hence water use per unit of leaf area [@Warren2010] leading to increased soil water availability. These ‘water savings’ can be expected to lead to an increase in *L* in water-limited ecosystems, consistent with a tight coupling of *L* with precipitation across gradients [@Ma2013; @Jin2013; @Hatton2008]. In support of this prediction, Donohue et al. [-@Donohue2013] attributed the observed global increase in *L* in arid and semi-arid areas over the last 30 years to rising C~a~. Based on the same water-savings principle, we predict that *L* should increase under eC~a~ in water-limited forests.

\
Most previous studies on *L* responses to environmental drivers have focused on the seasonal, mostly temperature and light-driven, timing of *L* dynamics observed in N-hemisphere forests [e.g. @Richardson2012; @Savoy2015] whereas *L* responses to fluctuations in water availability are qualitatively well-known [@Pook1985; @Specht1989; @Whitehead2004; @Pook1997; @Limousin2009], but poorly quantified. The strong relationship between *L* and rainfall observed across Australia [@Ellis2008; @Specht1989; @Broich2014], and other continents [@Donohue2013; @Grier1977; @Jin2013] represents the adaptation of vegetation structure and density to long-term climate conditions. However, local *L* is also highly responsive to water availability at shorter time scales [@Pook1985; @Pook1997], driven by the dynamics of leaf turnover [@Misson2011] and leaf flushing and expansion [@Muller2011], which are both highly responsive to soil water availability. Understanding dynamics of *L* in response to water availability and its potential interactions with C~a~ is crucial if we are to improve phenology models in water-limited ecosystems [@Misson2011].

\
We estimated *L* from near-continuous measurements of canopy transmittance in a *Eucalyptus*-dominated woodland exposed to eC~a~, the EucFACE experiment. The woodland at the EucFACE is frequently water-limited, allowing us to quantify the dynamics of *L*, and its components (litter fall and production) in response to water availability. We hypothesized that eC~a~ would result in an increase in *L*, particularly during periods of water limitation, as water savings were expected to lead to longer retention of foliage under elevated CO~2~. In addition, we expected that the low *L* that is typical of native *Eucalyptus*-dominated woodlands [@Whitehead2004] allowed a larger response to eC~a~ [cf. @Norby2011]. Our second objective was to quantify the temporal dynamics of *L*, in particular in relation to fluctuating water availability.

@@

## Site description and experimental design
The EucFACE experiment is located on an ancient alluvial floodplain about 5 km from the Hawkesbury River in western Sydney (Australia) at 30 m a.s.l. elevation (33^o^ 37'S, 150^o^ 44'E). It is located within a remnant patch  of native Cumberland Plain woodland [@Tozer2003]. The vegetation, maintained in this location since the 1880’s, retains some old-growth trees and some re-growth vegetation though it has been undisturbed for ca. 70 years. The site is characterized by a humid temperate-subtropical transitional climate with a mean annual temperature of 17 ^o^C and mean annual precipitation (P) of 800 mm (Bureau of Meteorology, station 067105 in Richmond, NSW Australia; http://www.bom.gov.au). The potential evapotranspiration (PET) is ca. 1500mm (*to be refined!*), giving a ratio of P/PET of ca. 0.53 (*quite dry - details to be added*).

\
The soil at the site is a loamy sand in the Clarendon Formation [@Barton2010; @Isbell2002], a slightly-acid, low-organic C, well-drained soil. The soil surface is a loamy sand with > 75% sand content, and from ca. 50 to 300 cm depth, soils are a sandy clay loam with > 30% silt and clay. The soil is nutrient poor, and a strong limitation of P on tree growth at the site has recently been demonstrated [@Crous2015]. Groundwater depth is 12.6 – 12.9 m below the soil surface. The woodland ranges in density from 600-1000 trees ha^-1^. The vegetation consists of an open woodland (basal area before treatment, BA = `r mean(ba$BA)`, SE = `r sd(ba$BA)/sqrt(6)` m^2^ha^-1^, n = 6 plots) with a single major canopy-forming tree (*Eucalyptus tereticornis* Sm.) with a minor component of *E. amplifolia* Naudin. The understory stratum is composed mainly of grass species dominated by *Microlaena stipoides* (Labill.) R.Br., with lower densities of forbs of *Commelina cyanea* R.Br., *Lobelia purpurascens* R.Br., *Dichondra repens* J.R.Forst. & G.Forst and occasional shrubs of *Breynia oblongifolia* (Mull.Arg.) Mull.Arg. and *Bursaria spinosa* Cav.

@@

where $\tau_p$ canopy transmission estimated from canopy photos. We assumed that the extinction coefficient *k* = 0.5, following Macfarlane et al. [-@Macfarlane2007] recommendation for eucalypt forests. Without further calibration, the two estimates were reasonably well correlated, though on average $L_\tau$ was `r 100*(with(flatcan_byring, mean(LAI.PAR.mean / LAI))-1)` higher than $L_p$ (Fig. SI2, $R^2$=`r glance(lmphoto)$r.squared`, RMSE=`r glance(lmphoto)$sigma`) (Fig. S2). We also estimated the extinction coefficient that minimized the difference between the two estimates of *L* (Fig. S2), giving *k* = `r kopt`. These comparisons are encouraging because the photos measure *L* near the zenith (+/- 15°), while $L_\tau$ is integrated over the sky hemisphere (Eq. 3), weighted towards the zenith.

@@

## Response of *L* to elevated [CO~2~]
We report on `r nummon` months of continuous leaf area index (*L*) measurements in a native evergreen *Eucalyptus* woodland exposed to +150ppm elevated CO~2~ (eC~a~), including 3 months of ramp-up in C~a~. We predicted that *L* would increase in response to eC~a~ as a result of enhanced CO~2~ assimilation for the entire canopy, and increased soil water availability. We also expected that this eucalypt woodland, with its inherently low *L* [@Whitehead2004], would be even more responsive to eC~a~ than higher-*L* and more closed-canopy stands [@Norby2011]. Counter to our expectations but similar to Norby et al. [-@Norby2003], we found no significant effect of elevated C~a~ on *L*. It could be argued that the change in *L* in response to eC~a~ was too small to be detected. Natural variation in *L* among plots in a native woodland is indeed a challenge to manipulative studies with low replication, such as FACE experiments. To increase sensitivity and statistical robustness, we used a novel method to account for pre-treatment differences by using stand basal area as a covariate. Using this method, we estimated that we can detect a difference of only a `r format(Ldetect,digits=1)`% in *L*. Based on this small detectable difference, we conclude that the lack of a significant C~a~ effect is not due to a lack of power, but a true lack of response to elevated C~a~.

\
This lack of a response to elevated C~a~ counters our expectations and those in previous publications [@Norby2003; @McCarthy2007], for a number of reasons. First, in a review of forest FACE sites, Norby and Zak [-@Norby2011] showed that low *L* stands were more responsive to eC~a~. They argued that high *L* canopies would not benefit from further increases in *L*, because it would lead to only very small increases in total absorbed radiation. The mean *L* at the EucFACE was ca. `r mean(ba$LAI.mean)` m^2^m^-2^, which is low compared to a global compendium of forest *L* [@Iio2013], and the lowest *L* of any forest FACE experiment [@Norby2011], but in line with a range of values for native *Eucalyptus* forests and woodlands reported by Whitehead and Beadle [-@Whitehead2004] and Iio et al. [-@Iio2013]. Based on this *L* estimate and the relationship across FACE sites reported by Norby and Zak [-@Norby2011], we could expect about a 30% increase in *L* under eC~a~. It is clearly too early to draw general conclusions across FACE sites in terms of *L* response, moreover because compare to other forest FACE sites, the EucFACE is very strongly limited by nutrient availability, in particular phosphorus [@Crous2015; @Nielsen2015].

\
A possible explanation for the lack of a *L* response to eC~a~ is that leaf growth may not be directly related to increased carbon availability as a result of eC~a~. In other words, stimulation of net leaf photosynthesis by eC~a~ (a common response, also observed at the EucFACE, Gimeno et al. *in review*) may not lead to increased net leaf growth, due to other constraints such as water or nutrient limitations [@Crous2015]. Net leaf growth is the net sum of allocation of photosynthates to leaf growth, and turnover of the leaf pool. Either process may itself be affected by eC~a~ [@DeKauwe2014]. We found a significant effect of eC~a~ on leaf lifespan, with leaves living approximately `r round(abs(LLtest$estimate*12),1)` months longer in elevated C~a~ (Fig. 6). Because we found no difference in *L* with C~a~ treatment, this suggests a compensating difference in the fractional allocation of net primary production to leaf growth. To further understand the response of biomass allocation to eC~a~ at the EucFACE, all components of the carbon balance must be carefully quantified, a task underway but not yet complete. For example, Drake et al. (*in review*) reported increased soil respiration under eC~a~, suggesting that at least some of the stimulation of photosynthesis was allocated to either root growth, respiration, or root exudation.

\
Despite a lack of response of canopy *L*, we did find a significantly longer leaf lifespan (LL) in eC~a~, consistent with our second hypothesis. We could not determine whether there was an interaction with water availability, since we estimated leaf turnover from mean *L* and total accumulated leaf litter (Fig. 6). Although LL was increased in eC~a~, this was not reflected in a difference in leaf litter production at any time point (Fig. 5), suggesting the change in LL reflects a small but significant shift in leaf retention, not related to decreased litter production during drought. There is little evidence from other FACE sites that LL is affected by eC~a~. Taylor et al. [-@Taylor2008] reported delayed autumn senescence in *Populus* in two FACE experiments (PopFACE and AspenFACE). Although this indicates the canopy stayed green longer in elevated C~a~, [@Gielen2001] reported faster turnover rate of leaves in elevated C~a~, leading to similar *L* in ambient C~a~ despite higher productivity. In contrast, Warren et al. [-@Warren2011] found increased senescence in eC~a~ at the ORNL FACE site (*Liquidambar styraciflua* plantation), in response to a combined drought and heat wave.

##Continuous measurement of *L*
Studying canopy leaf area index in an evergreen natural woodland poses unique challenges. Unlike in deciduous canopies, litter production cannot be used to estimate *L*, because leaf lifespan (LL) is unknown (instead, we used *L* and litter production to estimate LL). We can also not rely on leaf area estimates from allometric relationships with stem diameter or height, since these equations are not available for mature *E. tereticornis*, and would be useless anyway since *L* is highly variable temporally (Figs. 1, 7). We found that indirect estimates of *L* based on canopy transmittance were very reliable because relatively small temporal changes in *L* could be detected, indicating leaf flushing events after rainfall, increased litter production during drought, and an insect attack (Figs. 1, 5, 7). We do argue that the raw data must be filtered by diffuse conditions, as the spatial and within-day variance are negligible compared to periods of direct sunshine [@Chason1991; @Dufrene1995; @Nilson1999]. As expected, when we used unfiltered canopy transmittance data, the seasonal patterns in canopy transmittance were much more distinct (Fig. S3), because longer path lengths in the winter result in lower overall transmittance. Using diffuse PAR only avoids having to correct for these path length effects.

We used the opportunity provided by the high quality timeseries of *L* to test a simple photography-based method suggested by Macfarlane et al. [-@Macfarlane2007]. The two methods compared quite well (R^2^ = `r glance(lmcanphot)$r.squared`) (Fig. S2), suggesting the photography-based method can be reliably used in *Eucalyptus*-dominated woodlands to estimate *L*.

## Litter production and dynamics of *L*
We found a correlation between *L* dynamics and litter production, when the data were analyzed separately for periods of decreasing and increasing *L* (Fig. 4). As expected, during periods of decreasing *L*, the change in *L* was correlated with litter production, and we were able to use a four-month period in 2013 without significant rainfall to calibrate estimates of *L* against litter production. During periods of increasing *L*, litter fall increased sharply (Figs. 4 and 5), indicating replacement of old leaves by young leaves. Indeed we observed in the field that very few old leaves were present in the canopy after the significant flushing event of Dec. 2013. This is in line with Pook [-@Pook1984], who showed in *Corymbia maculata* (a genus closely related to *Eucalyptus*) accelerated litter production during periods of net *L* increase, but with a lag between litter fall and production of about 2 months. Similarly, Stone and Bacon [-@Stone1995] showed that drought-stressed trees had higher litter production at the branch scale, but irrigated trees had larger litter production at the canopy scale. Since leaf flushing has been shown to be closely related to soil water availability [Fig. 7; @Pook1984a], net leaf turnover is driven by water availability both through increased litter production during extended droughts, and complete canopy replacement following leaf flushing (Fig. 5). The consequence is that leaf turnover is highly variable in time (Fig. 5).

## Consequences for phenology models
Current-generation phenology models are heavily biased towards temperate conditions [@Primack2011], where leaf flushing occurs in spring in response to increase in temperature [@Richardson2012; @Savoy2015]. In contrast, phenology of evergreen *Eucalyptus* is more opportunistic, and shows rapid response to water availability (Fig. 7). Specht and Brouwer [-@Specht1975] demonstrated for four *Eucalyptus* species that leaf growth  peaked both in spring and autumn, with low winter temperatures and high summer temperatures (and low water availability) suppressing leaf growth. Similarly, Pook  [-@Pook1984a] demonstrated occurrence of leaf production during most months in *Corymbia maculata*, except during winter and suppression by dry periods. Thus a minimum temperature is also required for leaf growth to occur. We also have not observed leaf flushing in winter time, but at our site this was confounded with water availability as both winters during the study period were relatively dry (Fig. 7).

\
Although the response of *L* to water availability is qualitatively well-known, current-generation phenology models fall short in incorporating this response. In a comparison of six leading ecosystem and global vegetation models,  Powell et al. [-@Powell2013] showed that the steady decrease in *L* in an Amazon drought experiment could be matched by only one model. Simulations from De Kauwe et al. [-@DeKauwe2014] at ambient and elevated C~a~ also showed little response to inter-annual variation in rainfall (pers. comm. Martin De Kauwe). In most models  that provide prognostic estimates of *L* [i.e., excluding models that take *L* as an input parameter; e.g. @Kala2013], temporal changes in *L* are modelled by first calculating total net photosynthetic productivity (carbon availability, gross photosynthesis minus respiration), and allocating some fraction to the leaf pool. Combined with - usually constant - leaf turnover and specific leaf area, this yields predictions of *L* in response to environmental drivers such as eC~a~. We hypothesize that this algorithm is not well suited to explain the dynamic nature of *L* in response to water availability observed in evergreen broad-leaved forests. In order to construct a phenology model where phenology is responsive to water availability, it should include responses of both leaf turnover and leaf growth to fluctuating water availability, as we have demonstrated that both processes are highly dynamic. Water limitation is known to directly influence leaf turnover [@Pook1997], possibly to maintain sufficient rates of water availability to the remaining leaves [@Vilagrosa2003; @Misson2011], and rates of leaf expansion [e.g. @Muller2011]. Some existing ecosystem models allow for increased leaf turnover during periods of low water availability (e.g. G'DAY, see Corbeels et al. 2005), but this process has been poorly quantified.

# Conclusions
Leaf area index (*L*) of a natural *Eucalyptus* woodland was unresponsive to elevated [CO~2~] after `r nummon` months of treatment and `r nummon-3` months at +150ppm CO~2~ elevation. Litter production was closely related to dynamics of *L*, both during dry periods (leading to accelerated litter production), and periods of net leaf growth (when old foliage was replaced by new foliage). Leaf lifespan, estimated from mean *L* and annual litter production, was `r round(abs(LLtest$estimate*12),1)` months longer in the elevated [CO~2~] treatment.

\
We found no clear annual pattern in dynamics of *L*, instead, a net increase in *L*, indicating leaf growth, consistently commenced after heavy rainfall following a dry period. The complex dynamics of litter production and the rapid response of litter production and leaf growth to soil water availability challenge current-generation prognostic *L* and phenology models. High resolution time series such as the one we present in this study can be used as a basis to develop new models of phenology for evergreen woodlands.
